---
title: "Nanomia deseq"
format: html
editor: visual
---

## Set up

```{r}
#| output: false
library("DESeq2")
library(tidyverse)
library(apeglm)
library(pheatmap)
library(RColorBrewer)
```


```{r}
load("txi.RData")
```


Wrangle sample data.

```{r}
samples$collection_id <- gsub("-", "", samples$collection_id)

# Missing sample stages are mature
samples$stage[is.na(samples$stage)] <- "mature" 

# where stage="gonodendron", replace stage with stage_sex
samples$tissue = as.character(samples$tissue)
samples$tissue <- ifelse(!is.na(samples$sex), paste(samples$tissue, samples$sex, sep = "_"), samples$tissue)
samples$tissue <- as.factor(samples$tissue)
```

The male gonodendron sample is not replicated, and is an outlier in the PCA. Get the sample number so we can drop it from analyses.

```{r}
# to_drop = samples[samples$sex == "male",]$sample

to_drop = c(
	"SAMN11554597"#,     # Male gonodendron, very different from other samples
	#"SAMN11554609"      # This developing palpon does not fall out with other palpons
)
```

Building on [tutorial](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#differential-expression-analysis).

## QC

### Check for collection_id effect


```{r}
de_collection_tissue  <- DESeqDataSetFromTximport(txi,
	colData = samples,
	design = ~ collection_id + tissue )

# de_collection_tissue <- de_collection_tissue[, colnames(de_collection_tissue) %in% to_drop]

de_collection_tissue <- DESeq(de_collection_tissue)
res_collection_tissue <- results(de_collection_tissue)
```

```{r}
plotMA(de_collection_tissue, ylim=c(-2,2))
```

```{r}
summary(res_collection_tissue)
```

```{r}
rld <- DESeq2::rlog( de_collection_tissue ) #rlog transform
vsd <- DESeq2::varianceStabilizingTransformation( de_collection_tissue ) #variance Stabilizing Transformation
		
notAllZero <- ( rowSums( DESeq2::counts( de_collection_tissue ) )>0 )
```

```{r}
vsd_collection_tissue <- vst(de_collection_tissue, blind=FALSE)
```

```{r}
sampleDists <- dist(t(assay(vsd_collection_tissue)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd_collection_tissue$collection_id, vsd_collection_tissue$tissue, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

```{r}
pcaData <- plotPCA(vsd_collection_tissue, intgroup=c("collection_id", "tissue"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=tissue, shape=collection_id)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```


## tissue + stage patterns

May want to exclude the pneumatophore from some analyses.

The `results()` function provides an overview of pairwise comparisons for one or more factors.

```{r}
de_stage_tissue  <- DESeqDataSetFromTximport(txi,
	colData = samples,
	design = ~ stage + tissue )

de_stage_tissue <- DESeq(de_stage_tissue)
res <- results(de_stage_tissue)
```

```{r}
res <- results(de_stage_tissue)
summary(res)
```

```{r}
plotMA(res, ylim=c(-2,2))
```


```{r}
resultsNames(de_stage_tissue)
```

```{r}
resLFC_mature_vs_developing <- lfcShrink(de_stage_tissue, coef="stage_mature_vs_developing", type="apeglm")
resLFC_mature_vs_developing
plotMA(resLFC_mature_vs_developing, ylim=c(-2,2))
```

```{r}
ntd <- normTransform(de_stage_tissue)

select <- order(rowMeans(counts(de_stage_tissue,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(de_stage_tissue)[,c("stage","tissue")])

pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```


```{r}
vsd_stage_tissue <- vst(de_stage_tissue, blind=FALSE)
```

```{r}
sampleDists <- dist(t(assay(vsd_stage_tissue)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd_stage_tissue$stage, vsd_stage_tissue$tissue, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

```{r}
pcaData <- plotPCA(vsd_stage_tissue, intgroup=c("stage", "tissue"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=tissue, shape=stage)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

## Tissue and stage specific effects

```{r}

samples_sub = samples[! samples$sample %in% to_drop,]

# get a table of all unique combinations of tissue and stage in the dataset
tissue_stage <- as.data.frame(unique(samples_sub[,c("tissue","stage")]))

# Order it by tissue and stage
tissue_stage <- tissue_stage[order(tissue_stage$tissue, tissue_stage$stage),]
tissue_stage
```

```{r}

res_all = data.frame()
res_lfc_all = data.frame()

# loop over rows of tissue_stage
for (i in 1:nrow(tissue_stage)) {
	# Set up the focal tissue and stage
	tissue <- tissue_stage[i,]$tissue
	stage <- tissue_stage[i,]$stage
	treatment_name = paste0(tissue, "_", stage)
	print(treatment_name)
	samples$focal = c("other")
	samples$focal[samples$tissue == tissue & samples$stage == stage] = treatment_name
	samples$focal = factor(samples$focal)

	# Run DESeq2
	de  <- DESeqDataSetFromTximport(txi,
		colData = samples,
		design = ~ focal )

	de <- de[, ! colnames(de) %in% to_drop]

	de <- DESeq(de)

	# Run differential expression analysis
	res <- results(de, c("focal", treatment_name, "other"))
	summary(res)
	sum(res$padj < 0.1, na.rm=TRUE)

	# Create MA plot
	p_MA = plotMA(res, ylim=c(-2,2))
	pdf(paste0("MA_", treatment_name, ".plot.png"))
	p_MA
	dev.off()

	# collect the results
	df_res = as.data.frame(res)
	df_res$focal = c(treatment_name)
	res_all = rbind(res_all, df_res)

	# Run differential expression analysis with log fold change shrinkage
	rn = resultsNames(de)
	coef = rn[grep("focal", rn)[1]] # Get the fist element of rn that starts with "focal"
	resLFC <- lfcShrink(de, coef=coef, type="apeglm")
	summary(resLFC)

	# Create LFC MA plot
	p_MA_LFC =  plotMA(resLFC, ylim=c(-2,2))
	pdf(paste0("MA_LFC_", treatment_name, ".plot.png"))
	p_MA_LFC
	dev.off()

	# collect the results
	df_res_lfc = as.data.frame(resLFC)
	df_res_lfc$focal = c(treatment_name)
	res_lfc_all = rbind(res_lfc_all, df_res_lfc)

}
```

For now, just focus on genes with significant elevated expresssion.
```{r}
padj_thresh = 0.05
res_all = res_all[(res_all$padj < padj_thresh) & (res_all$log2FoldChange > 0),]
res_all = res_all[order(res_all$focal, -res_all$log2FoldChange),]

res_lfc_all = res_lfc_all[(res_lfc_all$padj < padj_thresh) & (res_lfc_all$log2FoldChange > 0),]
res_lfc_all = res_lfc_all[order(res_lfc_all$focal, -res_lfc_all$log2FoldChange),]
```


## Integrate annotations

```{r}
# Read in the annotations
annotations = read.csv("Nanomia_bijuga.pep.emapper.annotations", sep="\t", stringsAsFactors = FALSE, skip=4)
annotations <- rename(annotations, query = X.query)

res_all$gene = rownames(res_all)
res_lfc_all$gene = rownames(res_lfc_all)

res_all = left_join(res_all, annotations, by=join_by(gene==query))
res_lfc_all = left_join(res_lfc_all, annotations, by=join_by(gene==query))

```

```{r}
# Write out the results to tsv files
write.table(res_all, file="res_all.tsv", sep="\t", quote=FALSE, row.names=FALSE)
write.table(res_lfc_all, file="res_lfc_all.tsv", sep="\t", quote=FALSE, row.names=FALSE)
```


## Munro et al analyses

Previous analyses at: 

- https://github.com/dunnlab/Siphonophore_Expression/blob/master/Manuscript.Rmd

- https://github.com/dunnlab/Siphonophore_Expression

- https://academic.oup.com/mbe/article/39/2/msac027/6521037
