---
title: "Nanomia deseq"
format: html
editor: visual
---

## Set up

```{r}
#| output: false
library("DESeq2")
library(tidyverse)
library(apeglm)
library(pheatmap)
library(RColorBrewer)
```

```{r}
load("txi.RData")
```


```{r}

### the samples

```{r}
samples$collection_id <- gsub("-", "", samples$collection_id)
samples
```

The missing stages are mature.

```{r}
samples$stage[is.na(samples$stage)] <- "mature"
```

```{r}
# where stage="gonodendron", replace stage with stage_sex
samples$tissue = as.character(samples$tissue)
samples$tissue <- ifelse(!is.na(samples$sex), paste(samples$tissue, samples$sex, sep = "_"), samples$tissue)
samples$tissue <- as.factor(samples$tissue)
```

```{r}
```

```{r}
```

Building on [tutorial](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#differential-expression-analysis).

## QC

### Check for collection_id effect

This is what the Munro et al analyses were.

```{r}
de_id_tissue  <- DESeqDataSetFromTximport(txi,
	colData = samples,
	design = ~ collection_id + tissue)

de_id_tissue <- DESeq(de_id_tissue)
res_id_tissue <- results(de_id_tissue)
```

```{r}
plotMA1 <- function(ddsMF){
		  
			pairs <- combn(as.character(unique(ddsMF$tissue)),2)
			plotMA1<-list()
			for(i in seq(from=1, to=ncol(pairs))){
				resulttable <- DESeq2::results( 
				  ddsMF,
				  contrast=c( "tissue", as.character( pairs[,i] ) ), 
				  pAdjustMethod="bonferroni", alpha=0.05 )
				resulttable <- as.data.frame(resulttable)
				resulttable$padj[is.na(resulttable$padj)]<-1
				
				plotMA1[[i]] <- 
				  ggplot(resulttable) + 
				      geom_point(alpha=0.8,aes(x=log(baseMean),y=log2FoldChange,colour= padj < 0.05)) +
				      scale_colour_manual(values =c("TRUE"="red","FALSE"="grey40")) +
				      ggtitle(paste( as.character( pairs[,i]),collapse =" vs " ))
				plotMA1<<-plotMA1
			}
		}

plotMA1(de_id_tissue)
```

```{r}
rld <- DESeq2::rlog( de_id_tissue ) #rlog transform
vsd <- DESeq2::varianceStabilizingTransformation( de_id_tissue ) #variance Stabilizing Transformation
		
notAllZero <- ( rowSums( DESeq2::counts( de_id_tissue ) )>0 )
```

```{r}
#PCA plot
data <- DESeq2::plotPCA( rld, intgroup=c("tissue", "collection_id"), returnData=TRUE )
percentVar <- round( 100 * attr(data, "percentVar") )
		
pca <- 
  ggplot( data, aes( PC1, PC2, color=tissue, shape=collection_id ) ) +
  geom_point(size=3) + 
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +  
  ylab(paste0("PC2:",percentVar[2],"% variance"))

pca
```


## tissue + stage patterns

May want to exclude the pneumatophore from some analyses.

The `results()` function provides an overview of pairwise comparisons for one or more factors.

```{r}
de_stage_tissue  <- DESeqDataSetFromTximport(txi,
	colData = samples,
	design = ~ stage + tissue )

de_stage_tissue <- DESeq(de_stage_tissue)
res <- results(de_stage_tissue)
```

```{r}
res <- results(de_stage_tissue)
summary(res)
```


```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```


```{r}
plotMA(res, ylim=c(-2,2))
```


```{r}
```


```{r}
resultsNames(de_stage_tissue)
```

```{r}
resLFC_mature_vs_developing <- lfcShrink(de_stage_tissue, coef="stage_mature_vs_developing", type="apeglm")
resLFC_mature_vs_developing
plotMA(resLFC_mature_vs_developing, ylim=c(-2,2))
```

```{r}
ntd <- normTransform(de_stage_tissue)

select <- order(rowMeans(counts(de_stage_tissue,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(de_stage_tissue)[,c("stage","tissue")])

pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```


```{r}
vsd_stage_tissue <- vst(de_stage_tissue, blind=FALSE)


```

```{r}
sampleDists <- dist(t(assay(vsd_stage_tissue)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd_stage_tissue$stage, vsd_stage_tissue$tissue, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```


```{r}
plotPCA(vsd_stage_tissue, intgroup=c("stage", "tissue"))
```


```{r}
pcaData <- plotPCA(vsd_stage_tissue, intgroup=c("stage", "tissue"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=tissue, shape=stage)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

## Tissue and stage specific effects

```{r}
# get a table of all unique combinations of tissue and stage in the dataset
tissue_stage <- as.data.frame(unique(samples[,c("tissue","stage")]))

# Order it by tissue and stage
tissue_stage <- tissue_stage[order(tissue_stage$tissue, tissue_stage$stage),]
tissue_stage
```

```{r}

res_all = data.frame()
res_lfc_all = data.frame()

# loop over rows of tissue_stage
for (i in 1:nrow(tissue_stage)) {
	# Set up the focal tissue and stage
	tissue <- tissue_stage[i,]$tissue
	stage <- tissue_stage[i,]$stage
	treatment_name = paste0(tissue, "_", stage)
	print(treatment_name)
	samples$focal = c("other")
	samples$focal[samples$tissue == tissue & samples$stage == stage] = treatment_name
	samples$focal = factor(samples$focal)

	# Run DESeq2
	de  <- DESeqDataSetFromTximport(txi,
		colData = samples,
		design = ~ focal )

	de <- DESeq(de)

	# Run differential expression analysis
	res <- results(de, c("focal", treatment_name, "other"))
	summary(res)
	sum(res$padj < 0.1, na.rm=TRUE)

	# Create MA plot
	p_MA = plotMA(res, ylim=c(-2,2))
	pdf(paste0("MA_", treatment_name, ".plot.png"))
	p_MA
	dev.off()

	# collect the results
	df_res = as.data.frame(res)
	df_res$focal = c(treatment_name)
	res_all = rbind(res_all, df_res)

	# Run differential expression analysis with log fold change shrinkage
	rn = resultsNames(de)
	coef = rn[grep("focal", rn)[1]] # Get the fist element of rn that starts with "focal"
	resLFC <- lfcShrink(de, coef=coef, type="apeglm")
	summary(resLFC)

	# Create LFC MA plot
	p_MA_LFC =  plotMA(resLFC, ylim=c(-2,2))
	pdf(paste0("MA_LFC_", treatment_name, ".plot.png"))
	p_MA_LFC
	dev.off()

	# collect the results
	df_res_lfc = as.data.frame(resLFC)
	df_res_lfc$focal = c(treatment_name)
	res_lfc_all = rbind(res_lfc_all, df_res_lfc)

}
```

For now, just focus on genes with significant elevated expresssion.
```{r}
padj_thresh = 0.05
res_all = res_all[(res_all$padj < padj_thresh) && (res_all$log2FoldChange > 0),]
res_all = res_all[order(res_all$focal, -res_all$log2FoldChange),]

res_lfc_all = res_lfc_all[(res_lfc_all$padj < padj_thresh) && (res_lfc_all$log2FoldChange > 0),]
res_lfc_all = res_lfc_all[order(res_lfc_all$focal, -res_lfc_all$log2FoldChange),]
```

```{r}
```

```{r}
```


## Munro et al analyses

Reremembering how previous analyses worked here - https://github.com/dunnlab/Siphonophore_Expression/blob/master/Manuscript.Rmd

https://github.com/dunnlab/Siphonophore_Expression

https://academic.oup.com/mbe/article/39/2/msac027/6521037

This function creates the DESeq2 object

https://github.com/caseywdunn/agalmar/blob/e76bcb8180fecca5a7af9fe751b731b37ad265f4/R/utility_functions.R#L318

```{r eval=FALSE}
#' Creates a DESeqDataSet object
#'
#' @param object An Expression object
#' @param design a formula that explains the project design
#' @return A DESeqDataSet object
#' @export
setMethod( "create_DESeq2", signature( object = "Expression", design = "formula" ),
	function( object, design ) {

		colData = data.frame(
			treatment = object@treatment,
			individual = object@individual,
			row.names = object@library_id
		)

		dds = DESeq2::DESeqDataSetFromMatrix(
				countData = floor( object@x ),
				colData = colData,
				design = design
	        )

		return( dds[ rowSums( DESeq2::counts( dds ) ) > 1, ] )
	}
)
```

It is then called from here

https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/functions.R#L102

```{r eval=FALSE}
dgeresults <- function( object ){
	if( length( levels( object@individual ) ) >=2 ) {
		
		ddsMF <- agalmar::create_DESeq2( object, design = ~individual + treatment )
		
		###collapse technical replicates
		
		run=paste(ddsMF$individual,ddsMF$treatment,sep="_")
		
		if(length(unique(run))!=length(run)){
		ddsMF$run = ddsMF$run=as.numeric(as.factor(run))
		
		ddsMF <- DESeq2::collapseReplicates( ddsMF,ddsMF$run, renameCols=FALSE )
		}
		
		ddsMF <- DESeq2::DESeq( ddsMF )
		
		#plots
		
		plotMA1 <- function(ddsMF){
		  
			pairs <- combn(as.character(unique(ddsMF$treatment)),2)
			plotMA1<-list()
			for(i in seq(from=1, to=ncol(pairs))){
				resulttable <- DESeq2::results( ddsMF,contrast=c( "treatment", as.character( pairs[,i] ) ), pAdjustMethod="bonferroni", alpha=0.05 )
				resulttable <- as.data.frame(resulttable)
				resulttable$padj[is.na(resulttable$padj)]<-1
				plotMA1[[i]]<-ggplot(resulttable)+geom_point(alpha=0.8,aes(x=log(baseMean),y=log2FoldChange,colour= padj < 0.05))+scale_colour_manual(values =c("TRUE"="red","FALSE"="grey40"))+ggtitle(paste( as.character( pairs[,i]),collapse =" vs " ))
				plotMA1<<-plotMA1
			}
		}

	  plotMA1(ddsMF)
		
		#Transforming data
		rld <- DESeq2::rlog( ddsMF ) #rlog transform
		vsd <- DESeq2::varianceStabilizingTransformation( ddsMF ) #variance Stabilizing Transformation
		
		notAllZero <- ( rowSums( DESeq2::counts( ddsMF ) )>0 )

		#adding heatmap
		select <- order( rowMeans( counts( ddsMF,normalized=TRUE ) ), decreasing=TRUE )[1:20]
		nt <- DESeq2::normTransform( ddsMF )
		log2.norm.counts <- assay( nt )[ select, ]
		df <- as.data.frame( ddsMF@colData[ ,c( "treatment","individual" ) ] )
		
		count_heatmap <- pheatmap::pheatmap( log2.norm.counts, cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df )
		
		rld_heatmap<- pheatmap::pheatmap( assay( rld )[select,], cluster_rows=FALSE, show_rownames=FALSE,
													cluster_cols=FALSE, annotation_col=df )
		
		vsd_heatmap<- pheatmap::pheatmap( assay( vsd )[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df )
		
		#sample distances
		sampleDists <- stats::dist( t( assay( rld ) ) )
		
		sampleDistMatrix <- as.matrix( sampleDists )
		rownames( sampleDistMatrix ) <- paste( rld$treatment, rld$individual, sep="-" )
		
		colnames( sampleDistMatrix ) <- NULL
		colors <- grDevices::colorRampPalette( rev( brewer.pal( 9, "Greens" ) ) )( 255 )
		distance_heatmap<- pheatmap::pheatmap( sampleDistMatrix, clustering_distance_rows=sampleDists, clustering_distance_cols=sampleDists, col=colors )
		
		
		#PCA plot
		data <- DESeq2::plotPCA( rld, intgroup=c("treatment", "individual"), returnData=TRUE )
		percentVar <- round( 100 * attr(data, "percentVar") )
		
		pca <- ggplot( data, aes( PC1, PC2, color=individual, shape=treatment ) ) + scale_shape_manual(values=0:length(data$treatment)) + geom_point(size=3) + xlab(paste0("PC1: ",percentVar[1],"% variance")) +  ylab(paste0("PC2:",percentVar[2],"% variance"))
		
		setClass("gg")
		setClass(
		  Class="DGEplotresults",
		  representation = representation(
		    ddsMF="DESeqDataSet",
		    plotMA="list",
		    count_heatmap="pheatmap",
		    rld_heatmap="pheatmap",
		    vsd_heatmap="pheatmap",
		    distance_heatmap="pheatmap",
		    pca="gg",
		    rld="DESeqTransform",
		    vsd="DESeqTransform"
		  )
		)
		
		DGEPlots <- methods::new( "DGEplotresults" )
		  DGEPlots@ddsMF<-ddsMF
		  DGEPlots@plotMA<-plotMA1
		  DGEPlots@count_heatmap<-pheatmap( log2.norm.counts, cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df )
		  DGEPlots@rld_heatmap<-pheatmap( assay( rld )[select,], cluster_rows=FALSE, show_rownames=FALSE,cluster_cols=FALSE, annotation_col=df )
		  DGEPlots@vsd_heatmap<-pheatmap( assay( vsd )[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df )
		  DGEPlots@distance_heatmap<-pheatmap( sampleDistMatrix, clustering_distance_rows=sampleDists, clustering_distance_cols=sampleDists, col=colors )
		  DGEPlots@pca<-pca
		  DGEPlots@rld<-rld
		  DGEPlots@vsd<-vsd
		  
	 return(DGEPlots)
		
	}
}
```

And that function is called here:

https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/Manuscript_prep.R#L147

```{r eval=FALSE}
dge <- lapply( e, dgeresults )
```

Here is where the PCA is then run:\
\
https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/Manuscript.Rmd#L1868

```{r eval=FALSE}
nanomia_data<-DESeq2::plotPCA(dge$`HWI-ST625-51-C02UNACXX-7-NANOMIA`@rld, intgroup=c("treatment", "individual"), returnData=TRUE )
percentVar_nano <- round( 100 * attr(nanomia_data, "percentVar") )

nanomia_pca <- ggplot(nanomia_data, aes( PC1, PC2, color=individual, shape=treatment ) ) + scale_shape_manual(values=c("Gastrozooid developing"=15,"Gastrozooid mature"=0,"Pneumatophore"=8, "Nectophore developing"=19,"Nectophore developing "=18, "Palpons mature"=2,"Palpons developing"=17,"Gonodendron male"=13,"Gonodendron female"=10)) + geom_point(size=3) + xlab(paste0("PC1: ",percentVar_nano[1],"% variance")) +  ylab(paste0("PC2:",percentVar_nano[2],"% variance")) + guides(shape  = guide_legend(order = 1),color = guide_legend(order = 2))

```

Here is where the plot is generated:

https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/Manuscript.Rmd#L1871

```{r eval=FALSE}
nanomia_pca <- ggplot(nanomia_data, aes( PC1, PC2, color=individual, shape=treatment ) ) + scale_shape_manual(values=c("Gastrozooid developing"=15,"Gastrozooid mature"=0,"Pneumatophore"=8, "Nectophore developing"=19,"Nectophore developing "=18, "Palpons mature"=2,"Palpons developing"=17,"Gonodendron male"=13,"Gonodendron female"=10)) + geom_point(size=3) + xlab(paste0("PC1: ",percentVar_nano[1],"% variance")) +  ylab(paste0("PC2:",percentVar_nano[2],"% variance")) + guides(shape  = guide_legend(order = 1),color = guide_legend(order = 2))

```

And finally here is where it is rendered:

https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/Manuscript.Rmd#L1950

### Differential expression analyses

Within species analyses: which genes are specific to particular zooid types?

https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/Manuscript.Rmd#L617

Here is where the zooid expression is collated into a tibble:

https://github.com/dunnlab/Siphonophore_Expression/blob/f1808ad614413d325d9b025b4d06ef604070f5d3/Manuscript.Rmd#L984

And the tibble of expression unique to each zooid type is generated here:

unique_zooids<-unique_zooids %>% dplyr::left_join(blast_lookup,by="sequence_ids") %>% dplyr::left_join(GO,by="sequence_ids") %>% dplyr::left_join(reduced_nodes, by="sequence_ids")